A comparison of different methods for calculating tangent-stiffness matrices in a massively parallel computational peridynamics code.

### 2014-02-22 response/response.tex

{\color{red}
  \begin{itemize}
    \item 
        We have revised the brief description of peridynamics in page 9 to be more complete:
    \item
        Material from silling2007 is interpreted including: the definition of vector states,
        the derivation of conservation laws in terms of state-based peridynamics, the definition
        of the isotropic elastic solid AKA the Linear elastic Peridynamic Solid.
    \item
        It is the authors' opinion that the additional material mentioned above explains the equations involved
        in the peridynamic models for mechanics, used in relation to the peridynamics code
        studied in the work.

  \end{itemize}}
    \item 
        The manuscript seems to be written in a language more familiar to computer scientists. 
        I would encourage the authors to revise certain parts of the manuscript to be
        presented in a language more familiar to the computational mechanics community.

{\color{red}
  \begin{itemize}
    \item 
        We have revised certain parts of the manuscript for clarity, adding additional information for the reader's
        convenience:
    \item
        In subsection 'Implementing the complex-step method in Peridigm', explanation of what is meant by datatype,
        container class, template, scope and instantiation are included since these concepts may be unfamiliar to
        a computational mechanics audience at least as denoted by these terms. 
    \item 
        In subsection 'The Comparative Study', the way message passing computations can slow runtime is explained
        by an analogy to an experiment given as an exercise for the reader. 
    \item 
        In section 'Conclusions', the term byte-copiable is explained. 
    \item 
        It is the authors' opinion that the above revisions to the language of the manuscript allow the intended audience
        to follow the discussion where concepts from computer science are relevant.
  \end{itemize}}

{\color{red}
  \begin{itemize}
     \item 
         The authors have appended a more comprehensive literature review for the CS method to section 1.1.1.
         The aspects of CS investigated by the prior work, as well as any particularly interesting
         developments introduced by those works, have been listed. To argue for the novelty of the instant work,
         each prior work is compared to the instant work in terms of the the use of the CS method and the similarity of the fields of research of the works.
     \item
         The literature review concludes with an argument for the novelty of the instant
         work where the prior works are taken as a whole and compared to the instant work. It is the authors' opinion
         that the instant work is novel for at least the reasons set out in the literature review material, where
         the contribution to the CS method is in the comparison to other methods in terms of computational
         time,  parallel implementation, and implementation within \emph{Peridigm} .
  \end{itemize}}

  \item
      The comparisons of the different algorithms are only performed with respect to the
      construction of tangent-stiffness matrices. In particular, the authors always use the
      solution of AD at each time step to compute the entries of the tangent-stiffness
      matrices, even for the FD, CD, and CS methods. However, the authors need to
      demonstrate the performance of the CS method, in comparison to the other ones,
      with respect to the solution of real, ideally non-linear problems.

### 2014-02-22 jacobian_calc_CMAME.tex
spelling mistakes in manuscript fixed

\note[Michael]{re: I.4}
It is appropriate to detail the prior work discussing aspects of the
complex-step method in order to indicate what is new in the instant work.  The
CS method was introduced in a work by J. N. Lyness and C. B. Moler
\cite{lyness1967numerical}.  In this prior work, the CS method is identified in
an exact form and the comptuing of analytical derivatives is shown. This prior
work develops the CS method, while the instant work is about an application of
the CS method.

@@

A work by W. Squire and G. Trapp showns through empirical data that CS offers superior accuracy
to FD and CD \cite{squire1998using}. The instant work uses the CS method as shown in the
prior work and was motivated by the prior work's suggestion that CS is highly accurate. 
The instant work examines applications of the CS method beyond taking derivatives alone,
whereas the prior work does not.

A work by J. C. Newman, W. K. Anderson and L. D. Whitfield, discusses computing
sensitivities in a CFD analysis using complex step and compares the results to
central difference produced results \cite{newman1998}. The work is also notable
for identifying automatic differentiation as a competing method to CS and
suggesting that complex step could be used to compute Jacobians of residuals
for use in Newton-Kyrlov schemes for solving non-linear systems. The instant
work is novel with respect to the prior work because the instant work
investigates CS as applied to a numerical method for solving non-linear
systems, rather than as in the prior work investigating the derivatives
themselves produced with the CS method.

@@

Another work by A. Perez-Foguet, A. Rodriguez-Ferran, A. Huerta The prior work
applies CS in computing derivatives needed for integrating stress, obtaining
algorithmically consistent elastoplastic modulii, and obtaining the Jacobian of
the global residual, for FEM simulations using the the MRS-Lade material model
\cite{perez2012numerical}. In this prior work, CPU runtime is measured.
Comparing the prior work to the instant work, the prior work demonstrates CS on
a serial computer and discusses a classical solid mechanics formulation.
Automatic differentiation is not investigated.  Scalability studies with a
variety of equivalent, progressively more refined meshes are not done.
Scalability as a function of distributed system size is not investigated.  The
instant work is novel with respect to the prior work in offering a discussion
of CS as applied to peridynamics simulation on a parallel computer, and
comparison to automatic differentiation as a competing method.

A work by Joaquim R. R. A. Martins, Ilan M. Kroo and Juan J. Alonso
investigates using CS for computing aerodynamic sensitivities
\cite{martins2000automated}. Additionally higher order accuracy versions of CS
are discussed, as is the task of implementing complex-step in most widely used
general purpose programming languages. Also dicussed is the relative ease of
implementing a CS scheme versus an AD based scheme for computing sensitivities
in an established code. The instant work is novel with respect to the prior
work because the instant work investigates CS as applied to a numerical method
for solving non-linear systems, rather than as in the prior work investigating
the using the CS method to obtain sensitivities.

A work by W. K. Anderson, J. C. Newman, D. l. Whitfield and E. J. Nielsen discusses
the CS applied to computing sensitivities in CFD simulations \cite{anderson2001sensitivity}. CS is
compared to AD both in Fortran and C++ applications. In this prior work it is suggested that forward mode
AD is mathematically equivalent to first order CS. Forware mode AD isThe instant work is novel with respect to the prior
work because the instant work investigates CS as applied to a numerical method
for solving non-linear systems, rather than as in the prior work investigating
the derivatives themselves produced with the CS method.

@@

A work by  K. L. Lai and J. L. Crassidis discusses first and second order CS
methods for computing derivatives. Choosing optimal step size
\cite{lai2008extensions}. This prior work discusses computing first and second
order accurate derivatives with CS and choosing optimal step size.  The intant
work discusses using the derviatives computed with the CS method in a numerical
method for solving a non-linear system.

@@

A work by W. Jin, B. H. Dennis, and B. P. Wang discusses using CS to compute
sensitivities in a solid mechanics FEM analysis \cite{jin2010improved}. In the
prior work the derivitives produced by CS themselves are the object of
interest, where the instant work uses CS in a numerical method for solving a
non-linear system.

A work by A. Voorhees, H. Millwater, and R. Bagley, where H. Millwater is an author of
the instant work, dicusses the background of the CS method and frames it as a subset
of Fourier Differentiation. CS and Fourier Differentiation are then applied to 
computing sensitivities within a solid mechanics FEA \cite{voorhees2011complex}. A notable
technique demonstrated by this prior work is the representation of complex valued
nodal components as additional degrees of freedom in the FE model, to be interpreted
by a matrix representation of multicomplex numbers during computations, yet without storing
the entire matrix. In the prior work, CS is discussed and related to a broader class of differentiation methods and used
to compute sensitivities. In the instant work, CS is used in a numerical method for solving non-linear systems.

@@

The body of prior literature has introduced the application of CS to
single dimension and multidimensional derivatives, sensitivity analysis,
non-linear solution schemes, the extension of CS for higher accuracy, 
and implementation strategies for CS in new and existing codes.
With respect to the prior work dicussing the CS method, the primary novel
aspects of the instant work are then its dicussion of the application of CS to
peridynamics simulations run in \emph{Peridigm}, a parallel implementation of
CS and a compilation of CPU time data, where each of CS, AD, FD and CD are
compared.

@@

\subsection{Efficiency Data}
Efficiency, calculated as the average number of seconds taken per nonzero Jacobian matrix element, was plotted as a function of the number nonzero TS matrix elements. The single core results are shown in Fig.~\ref{fig:serial_efficiency}.  The ranking order of the methods
matched their speed ranking in Fig.~\ref{fig:serial_speed}. This plot indicates that efficiency is insensitive to the the number of nonzero TS elements.  
%
\begin{figure}[tbp]
  \centering
  \scalebox{1.0}{\input{./figs/serial_speed_rel.pgf}}
  \caption{Serial test series efficiency measurements.}
  \label{fig:serial_efficiency}
\end{figure}
%
The parallel efficiency results are shown in Fig.~\ref{fig:multi_efficiency}. The parallel efficiency plot shows that AD was slower than FD because for the given test problem it had a lower efficiency. Strangely AD appears to gain in efficiency over the last few data points in the plot. The reason behind this trend was not determined.
%
\begin{figure}[tbp]
  \centering
  \scalebox{1.0}{\input{./figs/multi_speed_rel.pgf}}
  \caption{Multicore test series efficiency measurements.}
  \label{fig:multi_efficiency}
\end{figure}

\note[Michael]{Re: I.5, everything after this line is changed}
\section{Convergence and displacement accuracy studies} 
\label{sec:PeridigmConvergenceStudy}
%
While the previous sections profiled the per-iteration performance of each of
the methods, it was necessary to conduct convergence rate studies in order to
get a complete picture. It is necessary for an analyist to know if the
accuracy differences measured in the per-iteration tests would amount to
differences in convergence rates among the methods if they were allowed to solve problems
normally. The question is posed, given that step-size $h$ is heuristically
selected by \emph{Perdigm} for the finite difference methods and presumable
makes a good choice, do the single-iteration Jacobian accuracy results predict
the ranking of the methods in terms of convergence rate for a given study? For
brevity, convergence rate was measured for each of the methods in solving just
the largest problem of the single-core test series. That problem was solved with each of the methods
both on a single core and all sixteen cores of a 'normal' node on \emph{Stampede}. As was seen in the per-iteration results,
Jacobian accuracy did not vary greatly from test to test, such that any pair of tests
from the two series could have been considered reasonably representative. As in
the validation problem shown in ~\ref{subsec:Validation}, would each of the methods return an
equal convergence rate so long as $h$ were optimally selected?

@@

The CS method was shown to be highly accurate in calculating tangent-stiffness
matrices by the standards of this study, and it is slower than other methods
for a naive implementation. In the context of future use of CS in
\emph{Peridigm}, CS holds the advantage of relying on byte-copyable data types,
while AD requires a complicated serialization and deserialization in order to
function as byte-copyable. This is fine for copying in between MPI processes,
but deserialization on an accelerator or GPU is very difficult, as this would
require compiling a version of the AD library being used that were compatible
with the accelerator or GPU being used. \note[Michael]{re:
I.3}\add[Michael]{For the reader's convenience, a collection of contiguously
stored data elements are said to be 'byte-copiable' when the the interval
in bytes of memory spanned by a data element in the collection is constant
throughout the collection. Suppose 'contiguously stored' means, stored in
memory with a minimum of unrelated infomation stored in between data
elements in the collection. Complicated objects, such as AD datatype
elements, do not have the same size in memory as other objects even of that
same class, for the reason that they themselves contain an uncertain
ammount of state data according to their design and work done in the
program. This means that a collection of such objects, even if they are
contiguously stored in memory, will not be alligned according to a constant
interval, such that elements within a collection cannot be located by
coordinates consisting of the location of the beginning of the first
element in the collection and an interval multiplied by a natural number.
This method of locating is integral to the serialized communication
implemented by MPI the libraries as well as CUDA, OpenCL, and others,
which allow only the transmission of collections of byte-copiable datatypes
(that is 'primitive datatypes'). A common method of bypassing this
limitation for communication purposes is to write the complicated data to a
'character buffer', and transmit the byte-copiable character buffer,
however this is useless unless the recipient device of the communication
can recover the original data and process it, which is impossible if the
recipient device only can work with a subset of the programming language
that the sending device must use, such as in a communication between a CPU
running C++ and a GPU running C (such as in CUDA programs), where the gap in
sophistication contains the features of the "complicated" datatype.} What this means is
that if \emph{Peridigm} or any simulation software were to be reformulated to
take advantage of accelerators or GPUs, the flexibilty of complex-step owed to
its simple implementation may make it more viable for that application given
its high accuracy. Additional performance gains may be achieved through the use
of the templated \emph{TPetra} package in \emph{Trilinos} which would allow the
use of complex data types natively, and alleviate the expense of the
dynamic-casts that were utilized in this study.  As a final comment, it should
be noted that from a code development perspective the CS method may be the
easiest of the methods to implement; It only requires one functional
evaluation, there is no heuristic choice of probe distance, and there is no
dependence on an external AD library.  From this perspective, the method may
still be an attractive method for TS matrix evaluation. \add[Michael]{However, at least for 
the models producing the results shown here, each of the
methods returned quadratic convergence, indicating each was sufficiently
accurate. 'Marching orders' for the analyst would be to select the swiftest
method, while it would seem accuracy above what is adequate does not enhance
the convergence rate or iteration speed until either classical Newton's convergence
conditions or the approximate case conditions are violated.}

>>>>

\note[Michael]{re: I.4}
It is appropriate to detail the prior work discussing aspects of the
complex-step method in order to indicate what is new in the instant work.  The
CS method was introduced in a work by J. N. Lyness and C. B. Moler
\cite{lyness1967numerical}.  In this prior work, the CS method is identified in
an series form and the computing of analytical derivatives is shown. This prior
work develops the CS method, while the instant work is about an application of
the CS method.

@@

A work by W. Squire and G. Trapp shows through empirical data that CS offers superior accuracy
to FD and CD \cite{squire1998using}. The instant work uses the CS method as shown in the
prior work and was motivated by the prior work's suggestion that CS is highly accurate. 
The instant work examines applications of the CS method beyond taking derivatives alone,
whereas the prior work does not.

A work by J. C. Newman, W. K. Anderson and L. D. Whitfield, discusses computing
sensitivities in a CFD analysis using complex step and compares the results to
central difference produced results \cite{newman1998}. The work is also notable
for identifying automatic differentiation as a competing method to CS and
suggesting that complex step could be used to compute Jacobians of residuals
for use in Newton-Krylov schemes for solving non-linear systems. The instant
work is novel with respect to the prior work because the instant work
investigates CS as applied to a numerical method for solving non-linear
systems, rather than as in the prior work investigating the derivatives
themselves produced with the CS method.

@@

Another work by A. Perez-Foguet, A. Rodriguez-Ferran, A. Huerta The prior work
applies CS in computing derivatives needed for integrating stress, obtaining
algorithmically consistent elastoplastic modulii, and obtaining the Jacobian of
the global residual, for FEM simulations using the MRS-Lade material model
\cite{perez2012numerical}. In this prior work, CPU runtime is measured.
Comparing the prior work to the instant work, the prior work demonstrates CS on
a serial computer and discusses a classical solid mechanics formulation.
Automatic differentiation is not investigated.  Scalability studies with a
variety of equivalent, progressively more refined meshes are not done.
Scalability as a function of distributed system size is not investigated.  The
instant work is novel with respect to the prior work in offering a discussion
of CS as applied to peridynamics simulation on a parallel computer, and
comparison to automatic differentiation as a competing method.

A work by Joaquim R. R. A. Martins, Ilan M. Kroo and Juan J. Alonso
investigates using CS for computing aerodynamic sensitivities
\cite{martins2000automated}. Additionally higher order accuracy versions of CS
are discussed, as is the task of implementing complex-step in most widely used
general purpose programming languages. Also discussed is the relative ease of
implementing a CS scheme versus an AD based scheme for computing sensitivities
in an established code. The instant work is novel with respect to the prior
work because the instant work investigates CS as applied to a numerical method
for solving non-linear systems, rather than as in the prior work investigating
the using the CS method to obtain sensitivities.

A work by W. K. Anderson, J. C. Newman, D. l. Whitfield and E. J. Nielsen discusses
the CS applied to computing sensitivities in CFD simulations \cite{anderson2001sensitivity}. CS is
compared to AD both in Fortran and C++ applications. In this prior work it is suggested that forward mode
AD is mathematically equivalent to first order CS. The instant work is novel with respect to the prior
work because the instant work investigates CS as applied to a numerical method
for solving non-linear systems, rather than as in the prior work investigating
the derivatives themselves produced with the CS method.

@@

A work by  K. L. Lai and J. L. Crassidis discusses first and second order CS
methods for computing derivatives. Choosing optimal step size
\cite{lai2008extensions}. This prior work discusses computing first and second
order accurate derivatives with CS and choosing optimal step size.  The instant
work discusses using the derivatives computed with the CS method in a numerical
method for solving a non-linear system.

@@

A work by W. Jin, B. H. Dennis, and B. P. Wang discusses using CS to compute
sensitivities in a solid mechanics FEM analysis \cite{jin2010improved}. In the
prior work the derivatives produced by CS themselves are the object of
interest, where the instant work uses CS in a numerical method for solving a
non-linear system.

A work by A. Voorhees, H. Millwater, and R. Bagley, where H. Millwater is an author of
the instant work, discusses the background of the CS method and frames it as a subset
of Fourier Differentiation. CS and Fourier Differentiation are then applied to 
computing sensitivities within a solid mechanics FEA \cite{voorhees2011complex}. A notable
technique demonstrated by this prior work is the representation of complex valued
nodal components as additional degrees of freedom in the FE model, to be interpreted
by a matrix representation of complex numbers during computations, yet without storing
the entire matrix. In the prior work, CS is discussed and related to a broader class of differentiation methods and used
to compute sensitivities. In the instant work, CS is used in a numerical method for solving non-linear systems.

@@

The body of prior literature has introduced the application of CS to
single dimension and multidimensional derivatives, sensitivity analysis,
non-linear solution schemes, the extension of CS for higher accuracy, 
and implementation strategies for CS in new and existing codes.
With respect to the prior work discussing the CS method, the primary novel
aspects of the instant work are then its discussion of the application of CS to
peridynamics simulations run in \emph{Peridigm}, a parallel implementation of
CS and a compilation of CPU time data, where each of CS, AD, FD and CD are
compared.

@@

\subsection{Efficiency Data}
Efficiency, calculated as the average number of seconds taken per nonzero Jacobian matrix element, was plotted as a function of the number nonzero TS matrix elements. The single core results are shown in Fig.~\ref{fig:serial_efficiency}.  The ranking order of the methods
matched their speed ranking in Fig.~\ref{fig:serial_speed}. This plot indicates that efficiency is insensitive to the number of nonzero TS elements.  
%
\begin{figure}[tbp]
  \centering
  \scalebox{1.0}{\input{./figs/serial_speed_rel.pgf}}
  \caption{Serial test series efficiency measurements.}
  \label{fig:serial_efficiency}
\end{figure}
%
The parallel efficiency results are shown in Fig.~\ref{fig:multi_efficiency}. The parallel efficiency plot shows that AD was slower than FD because for the given test problem it had a lower efficiency. Strangely AD appears to gain in efficiency over the last few data points in the plot. The reason behind this trend was not determined.
%
\begin{figure}[tbp]
  \centering
  \scalebox{1.0}{\input{./figs/multi_speed_rel.pgf}}
  \caption{Multicore test series efficiency measurements.}
  \label{fig:multi_efficiency}
\end{figure}

\note[Michael]{Re: I.5, everything after this line is changed}
\section{Convergence and displacement accuracy studies} 
\label{sec:PeridigmConvergenceStudy}
%
While the previous sections profiled the per-iteration performance of each of
the methods, it was necessary to conduct convergence rate studies in order to
get a complete picture. It is necessary for an analyst to know if the
accuracy differences measured in the per-iteration tests would amount to
differences in convergence rates among the methods if they were allowed to solve problems
normally. The question is posed, given that step-size $h$ is heuristically
selected by \emph{Perdigm} for the finite difference methods and presumable
makes a good choice, do the single-iteration Jacobian accuracy results predict
the ranking of the methods in terms of convergence rate for a given study? For
brevity, convergence rate was measured for each of the methods in solving just
the largest problem of the single-core test series. That problem was solved with each of the methods
both on a single core and all sixteen cores of a 'normal' node on \emph{Stampede}. As was seen in the per-iteration results,
Jacobian accuracy did not vary greatly from test to test, such that any pair of tests
from the two series could have been considered reasonably representative. As in
the validation problem shown in ~\ref{subsec:Validation}, would each of the methods return an
equal convergence rate so long as $h$ were optimally selected?

@@

The CS method was shown to be highly accurate in calculating tangent-stiffness
matrices by the standards of this study, and it is slower than other methods
for a naive implementation. In the context of future use of CS in
\emph{Peridigm}, CS holds the advantage of relying on byte-copyable data types,
while AD requires a complicated serialization and deserialization in order to
function as byte-copyable. This is fine for copying in between MPI processes,
but deserialization on an accelerator or GPU is very difficult, as this would
require compiling a version of the AD library being used that were compatible
with the accelerator or GPU being used. \note[Michael]{re:
I.3}\add[Michael]{For the reader's convenience, a collection of contiguously
stored data elements are said to be 'byte-copiable' when the interval
in bytes of memory spanned by a data element in the collection is constant
throughout the collection. Suppose 'contiguously stored' means, stored in
memory with a minimum of unrelated information stored in between data
elements in the collection. Complicated objects, such as AD datatype
elements, do not have the same size in memory as other objects even of that
same class, for the reason that they themselves contain an uncertain
amount of state data according to their design and work done in the
program. This means that a collection of such objects, even if they are
contiguously stored in memory, will not be alligned according to a constant
interval, such that elements within a collection cannot be located by
coordinates consisting of the location of the beginning of the first
element in the collection and an interval multiplied by a natural number.
This method of locating is integral to the serialized communication
implemented by MPI the libraries as well as CUDA, OpenCL, and others,
which allow only the transmission of collections of byte-copiable datatypes
(that is 'primitive datatypes'). A common method of bypassing this
limitation for communication purposes is to write the complicated data to a
'character buffer', and transmit the byte-copiable character buffer,
however this is useless unless the recipient device of the communication
can recover the original data and process it, which is impossible if the
recipient device only can work with a subset of the programming language
that the sending device must use, such as in a communication between a CPU
running C++ and a GPU running C (such as in CUDA programs), where the gap in
sophistication contains the features of the "complicated" datatype.} What this means is
that if \emph{Peridigm} or any simulation software were to be reformulated to
take advantage of accelerators or GPUs, the flexibility of complex-step owed to
its simple implementation may make it more viable for that application given
its high accuracy. Additional performance gains may be achieved through the use
of the templated \emph{TPetra} package in \emph{Trilinos} which would allow the
use of complex data types natively, and alleviate the expense of the
dynamic-casts that were utilized in this study.  As a final comment, it should
be noted that from a code development perspective the CS method may be the
easiest of the methods to implement; It only requires one functional
evaluation, there is no heuristic choice of probe distance, and there is no
dependence on an external AD library.  From this perspective, the method may
still be an attractive method for TS matrix evaluation. \add[Michael]{However, at least for 
the models producing the results shown here, each of the
methods returned quadratic convergence, indicating each was sufficiently
accurate. 'Marching orders' for the analyst would be to select the swiftest
method, while it would seem accuracy above what is adequate does not enhance
the convergence rate or iteration speed until either classical Newton's convergence
conditions or the approximate case conditions are violated.}
